Controlled Experiments for Word Embeddings

### 2015-09-03 article/main.tex
incorporate the changes by adriaan

% compile using latex (xelatex seems to crop the figure strangely)
\documentclass{article} % For LaTeX2e
\usepackage[dvips]{graphicx}
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{mdframed}
%\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[outdir=./]{epstopdf}

\usepackage{graphics,graphicx}
\usepackage{amsmath,amssymb,amsfonts,comment}
%\definecolor{LinkColor}{rgb}{0,0,0}

\newcommand{\p}{\text{P}}
\newcommand{\word}[1]{\texttt{#1}}
\urldef{\googlegroup}\url{https://groups.google.com/forum/#!forum/word2vec-toolkit}
\urldef{\blogpost}\url{https://blog.lateral.io/2015/06/the-unknown-perils-of-mining-wikipedia/}

\title{Corpus Experiments for Word Embeddings}

 \author{
 	Benjamin Wilson\\
	Lateral GmbH\\
	\texttt{benjamin@lateral.io}
	\And
	Adriaan Schakel\\
	\textcolor{red}{NNLP?}\\
	\texttt{adriaan.schakel@gmail.com}
 }

@@

\begin{abstract}
	We introduce an experimental approach to studying the properties of word embeddings.
	Controlled experiments, achieved through modifications of the training corpus, permit the demonstration of direct, causal relationships between word properties and word vector direction and length.
	This approach is demonstrated in the case of the word2vec CBOW model with experiments that independently vary word frequency and word co-occurrence noise.
	In this case, it is shown that word vector length depends directly and linearly on both word frequency and the level of noise in the co-occurrence distribution of the word.
	In both cases, the coefficient of linearity depends upon the word.
\end{abstract}

\begin{section}{Introduction}
Word embeddings are typically trained with samples from the word co-occurrence distributions.
Recall that the co-occurrence distribution of a word $w$ gives the probability $\p(w'|w)$ that a word $w'$ occurs nearby, given that $w$ occurred.
Samples of the co-occurrence distributions are obtained by scanning a short window over the text.
The neighbouring words are treated as samples of the co-occurrence distribution the current word. 
Viewed in this way, the word vectors of the word embedding are determined by the co-occurrence distributions and frequencies of the words in the corpus.

We perform experiments that illustrate the effect of these two determining pieces of information (the co-occurrence distribution and the word frequency) as they are varied independently of one another.
This is achieved through the introduction of new tokens into the training corpus with varying frequencies and varying levels of noise in their co-occurrence distributions.
The frequency and co-occurrence distributions of these introduced tokens are modeled on existing words in the corpus.

We illustrate our approach in case of the word2vec CBOW model.
However, these experiments (and other besides) could equally be carried out for word embedding methods such as word2vec skip-gram \cite{DistRepns,EfficientEstimation}, GloVe \cite{pennington2014glove} and SENNA \cite{collobert-2011}.

In the case of word2vec, we show that word vector length depends directly and linearly on both word-frequency and the level of noise in the co-occurrence distribution of the word.
In both cases, the coefficient of linearity depends upon the word.
If the co-occurrence distribution is fixed, then word vector length increases with word frequency.
On the other hand, if word frequency is held constant, then word vector length decreases as the level of noise in the co-occurrence distribution of the word is increased.

We show furthermore that word vector direction does not vary with word frequency or with the level of co-occurrence noise.

This paper is structured as follows.
Section \ref{related-work} draws connections to related work, while section \ref{corpus-and-model} describes the corpus and model used to illustrate the experiments.
Section \ref{WFVE} describes a controlled experiment for varying word frequency while holding the co-occurrence distribution fixed.
Section \ref{CNVE}, in a complementary fashion, describes a controlled experiment for varying the level of noise in the co-occurrence distribution of a word, while holding the word frequency fixed.
Section \ref{future-directions} considers further questions and possible future directions.
\end{section}

\begin{section}{Related work}\label{related-work}
Our experimental finding (in the context of word2vec), that word vector length decreases with co-occurrence noise is reminiscent of \cite{vecchi-baroni-zamparelli2011}, where a relationship between vector length and the ``semantic deviance'' of an adjective-noun composition was studied observationally.
In \cite{schakel-wilson}, the interpretation of word vector length as an indication of word significance is explored in the context of the arXiv high-energy physics corpus.

\textcolor{red}{what about some work introducing tokens into a corpus for experiments?}

@@

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia datadump from October 2013.
In order to remove from the training data the bulk of robot-generated pages on Wikipedia, only pages with at least 20 monthly page views are retained\footnote{For further justification of this rationale, and to obtain the dataset, see \blogpost}.
Stubs and disambiguation pages are also removed, leaving 463 thousand pages with a total of 482 million words.
Word frequency is described in overview in Table \ref{word-occurrences-table}.

This base corpus is then modified as described in sections \ref{WFVE} and \ref{CNVE}.
Punctuation and numbers were removed and the corpus lower-cased prior to the modifications.
For recognisability, the tokens introduced into the corpus during the modification are uppercased.

For simplicity, only the word2vec CBOW word embedding with a single set of hyperparameters is considered.
Specifically, a $100$-dimensional CBOW model is trained using negative sampling with 5 negative samples, a window size of $10$, a minimum word occurrence of $128$, and $10$ passes through the corpus.
Sub-sampling was not used so that the influence of word frequency could be more clearly discerned.
Identical experimental results were obtained using hierarchical softmax, but these are omitted for succinctness.
The relatively high minimum word cut-off is chosen to ensure that word vectors, in all but degenerate cases, receive a sufficient number of gradient updates to be meaningful.
This cut-off results in a vocabulary of $81117$ words (only unigrams were considered).

The most recent revision of word2vec was used\footnote{SVN revision 42, see \url{http://word2vec.googlecode.com/svn/trunk/}}.
The source code for performing the experiments is made available on GitHub\footnote{\url{https://github.com/benjaminwilson/word2vec-norm-experiments}}.

Word vectors are then derived from the first layer of synaptic weights (``syn0'') and the second layer of weights (``syn1neg'') are discarded.

\begin{table}\label{word-occurrences-table}
	\input{words-occurrences.tex}
	\caption{ Number of words, by frequency band, as observed in the unmodified corpus.  }
\end{table}

\begin{subsection}{Replacement procedure}\label{replacement-procedure}
In the experiments, tokens are introduced into corpus via a replacement procedure.
To illustrate this procedure, consider a word, e.g. \word{cat}.
For each occurrence of this word, a sample $i$, $1 \leqslant i \leqslant n$ is drawn from a truncated geometric distribution, and that occurrence of \word{cat} is replaced with \word{CAT\_i}.
Thus the word \word{cat} is replaced throughout the corpus by a family of tokens with varying frequencies but approximately the same co-occurrence distribution as \word{cat}.

The geometric distribution is truncated in order to limit the number of tokens introduced into the corpus.
For any ratio $0 < \lambda < 1$ and maximum value $n > 0$, the truncated geometric distribution is given by
$$ P_{\lambda, n} (i) = \frac{\lambda^{i-1} \cdot (1-\lambda)}{(1 - \lambda^n)}, \qquad 1 \leqslant i \leqslant n.$$ 
This is the distribution over $i = 1, 2, \dots n$ for which the probabilities decay exponentially base $\lambda$.
Of course, other distributions might equally have been chosen for the experiments.
\end{subsection}
\end{section}

\begin{section}{Word frequency variation experiment}\label{WFVE}
In this experiment, families of tokens are introduced into the corpus.
The tokens in each family vary in frequency but share a common co-occurrence distribution.
After model training, comparison of the resulting word vectors associated to these tokens reveals the independent effect of frequency variation on the word embedding.

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
A small set of words from the unmodified corpus is chosen uniformly at random from the vocabulary.
In order that the introduced tokens do not have too low a frequency, only words which occur at least 10 thousand times are chosen.
To this set of words, the word \word{the} is added, in order to include a high-frequency stopword.
Table \ref{word-frequency-counts} lists the words chosen for this experiment along with their frequencies in the corpus.

The replacement procedure of subsection \ref{replacement-procedure} is then performed for each of these words, using a geometric rate of decay of $\lambda = 1/2$, and maximum value $n=20$.
This value of $\lambda$ is one of a range of values that ensure that, for each word, multiple tokens will be introduced with a number of occurrences sufficient to survive the chosen word2vec minimum occurrence cut-off of 128.  
A maximum value $n=20$ suffices for this choice of $\lambda$, since $2^{20 + \log_2{128}}$ exceeds the maximum occurrence of any token in the corpus. 
Figure \ref{fig:word-frequency-experiment-text-cat} illustrates the effect of these modifications on a sample text, with a family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that the word \word{cat} has been replaced throughout with the tokens \word{CAT\_i}.

\begin{table}\label{word-frequency-counts}
	\input{word-frequency-experiment-counts.tex}
	\caption{Occurrence counts for words chosen for the word frequency experiment. }
\end{table}

\begin{figure}\label{fig:word-frequency-experiment-text-cat}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{A text modified for the word frequency experiment as per subsection \ref{WFVEexisting}, where the
	word \word{cat} was chosen, $\lambda=0.5$ and $n=20$.}
\end{figure}

@@

\begin{subsection}{Tokens derived from an introduced, meaningless word}\label{WFVEmeaningless}
	A high-frequency, purely meaningless word is included in the experiment for comparison.
	We choose to introduce a new, entirely meaningless token \word{VOID} into the corpus, rather than choose an existing word whose meaninglessness is only supposed.
	To achieve this, the token is interspersed uniformly at random throughout the corpus so that its global frequency is $f = 0.005$.
	The co-occurrence distribution of \word{VOID} thus coincides with the global frequency distribution.
	The replacement procedure is then performed for \word{VOID}, using the same values for $\lambda$ and $n$ as above.
	figure \ref{fig:word-frequency-experiment-text-void} shows the effect of this modifications on a sample text ($f = 0.05$ is used there for illustrative purposes).

\begin{figure}\label{fig:word-frequency-experiment-text-void}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{A text modified for the word frequency experiment as per subsection \ref{WFVEmeaningless}, where $\lambda=0.5$ and $n=20$. For illustrative purposes, the token \word{VOID} was here interspersed with a global frequency of $0.05$.}
\end{figure}
\end{subsection}

\subsection{Results}
\textcolor{red}{rework}
For well-learned representations, word frequency has no effect on word vector direction.
Figure \ref{word-frequency-experiment-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus.
Recall that the cosine similarity measures the extent to which two vectors have the same direction, taking a maximum value of $1$ and a minimum value of $-1$.
Notice that the word vectors associated to tokens derived from a word other that \word{the} and \word{VOID} have the same direction.
The same is true of the word vectors associated to the \word{VOID\_i} for $i$ \textcolor{red}{below $7$ or $8$}, but that the word vectors \word{VOID\_i} gradually diverge directionally for higher values of $i$.
This demonstrates the difficulty of learning the co-occurrence distribution of $\word{VOID}$, the global frequency distribution.
Because of its noisiness, a higher number of samples (occurrences) is required to learn this co-occurrence distribution.
The same holds, but to a less pronounced extent, for the co-occurrence distribution of the stopword \word{the}.
\textcolor{red}{The poorly learned representations for \word{the\_i} for $i = 15, \dots $ and for \word{VOID\_i} for $i = 7, \dots$ are therefore disregarded.}

\textcolor{red}{include a table with nearest neighbours in the normal vocab for the to support this argument}
Table \ref{word-frequency-experiment-cosine-similarity}

It remains, therefore, to consider the effect of frequency variation on word vector length.
Figure \ref{fig:word-frequency-experiment-graph}, on the other hand, shows this relationship for individual words, both for the word vectors, ``syn0'' and for the discarded second layer vectors ``syn1neg'' (included for completeness).
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
For example, the six points on the line corresponding to \word{protestant} are labeled, from right to left, by the tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots, \word{PROTESTANT\_6}.
The number of points on the line is determined by the frequency of the original word.
For example, the frequency of the \word{protestant} can be halved at most $6$ times and remain about the minimum frequency cut-off.

Because all the points on a line share a co-occurrence distribution (only the frequency varies), figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively that length does indeed depend on frequency directly, and not only via some correlation between word frequency and word significance.
Moreover, this relationship is linear for any particular word.
Notice also that the relative positions of the norms of the word vectors associated to the test words is independent of the frequency band.
Therefore these lines trace lines along which a frequency-adjusted norm would be constant.

Notice also that the word vector for \word{VOID} is non-zero.

\begin{figure}\label{word-frequency-experiment-heatmap}
	\includegraphics[scale=0.5]{word-frequency-experiment-heatmap}
	\caption{
	Heatmap of the cosine similarity of the vectors associated to the
	tokens introduced into the corpus in the co-occurrence noise experiment
	(words other than \word{the} and \word{VOID} chosen randomly).
	}
\end{figure}

\begin{table}\label{word-frequency-experiment-cosine-similarity}
	\input{cosine-similarity.tex}
	\caption{
		The most similar words in the original vocabulary to the tokens
		introduced into the corpus associated to the word \word{the},
		and their cosine similarity with the most frequent such token,
		\word{THE\_1}.  It is apparent from the nearest neigbours that
		the word vectors of the low frequency tokens have not been
		adequately trained (there were inadequate samples).
	}
\end{table}

\begin{sidewaysfigure*}\label{fig:word-frequency-experiment-graph}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{
	Vector length vs frequency for the words chosen for the word frequency
	experiment.  For each word, tokens of varying frequency but with the
	co-occurrence distribution of that word were introduced into the
	corpus, as described in section \ref{WFVE}.
	The word vectors are obtained from the first synaptic layer, ``syn0''.
	The second layer ``syn1neg'' is included for completeness.
	}
\end{sidewaysfigure*}

\begin{section}{Co-occurrence noise variation experiment}\label{CNVE}
Here, and throughout, the noise distribution is taken to be the global word frequency distribution.
Thus noise can be added to the co-occurrence distribution of a word by interspersing occurrences of that word uniformly at randomly throughout the corpus.
A small set of words is chosen from the unmodified corpus in the same manner as in section \ref{WFVE}.

>>>>

% compile using latex (xelatex seems to crop the figure strangely)
\documentclass{article} % For LaTeX2e
\usepackage[dvips]{graphicx}
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{mdframed}
\usepackage{rotating}
\usepackage[outdir=./]{epstopdf}

\usepackage{graphics,graphicx}
\usepackage{amsmath,amssymb,amsfonts,comment}

\newcommand{\p}{\mathrm{P}}
\newcommand{\word}[1]{\texttt{#1}}
\newcommand{\mar}[1]{\marginpar{\footnotesize \bfseries
    \raggedright #1}}

\urldef{\googlegroup}\url{https://groups.google.com/forum/#!forum/word2vec-toolkit}
\urldef{\blogpost}\url{https://blog.lateral.io/2015/06/the-unknown-perils-of-mining-wikipedia/}

\title{Controlled Experiments for Word Embeddings}

 \author{
 	Benjamin Wilson\\
	Lateral GmbH\\
	\texttt{benjamin@lateral.io}
	\And
	Adriaan Schakel\\
	NNLP\\
	\texttt{adriaan.schakel@gmail.com}
 }

@@

\begin{abstract}
	An experimental approach to studying the properties of word
        embeddings is proposed.  Controlled experiments, achieved
        through modifications of the training corpus, permit the
        demonstration of direct relations between word properties and
        word vector direction and length.  The new approach is
        demonstrated using the word2vec CBOW model with experiments that
        independently vary word frequency and word co-occurrence noise.
        The experiments reveal that word vector length depends more or
        less linearly on both word frequency and the level of noise in
        the co-occurrence distribution of the word.  The coefficients of
        linearity depend upon the word.
\end{abstract}

\begin{section}{Introduction}
The field of natural language processing (NLP) is presently undergoing rapid
developments, not in the least triggered by the tremendous commercial
promises it holds.  A timely research topic in NLP are word embeddings,
where words from a vocabulary are represented by dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense
vectors of dimension much smaller than the vocabulary size are
constructed such that they carry syntactic and semantic information.
Irrespective of the technique chosen, word embeddings are
typically derived from word co-occurrences.  More specifically, in a
machine-learning setting, word embeddings are typically trained by
scanning a short window over all the text in a corpus.  This process can
be seen as sampling word co-occurrence distributions, where it is
recalled that the co-occurrence distribution of a target word $w$
denotes the conditional probability $\p(w'|w)$ that a word $w'$ occurs in
its context, i.e., given that $w$ occurred.  Most applications of word
embeddings explore not the word vectors themselves, but relations
between them to solve, for example, similarity and word relation tasks
\cite{vecchi-baroni-zamparelli2011}.  For these tasks, it was found that using normalised word
vectors improves performance.  Word vector length is therefore typically
ignored.

In a previous paper \cite{schakel-wilson}, we proposed the use of word
vector length as measure of word significance.  Using a domain-specific
corpus of scientific abstracts, we observed that words that appear only
in similar contexts tend to have longer vectors than words of the same
frequency that appear in a wide variety of contexts.  For a given
frequency band, we found meaningless function words clearly separated
from proper nouns, each of which typically carries the meaning of a
distinctive context in this corpus.  In other words, the longer its
vector, the more significant a word is.  We also observed that word
significance is not the only factor determining the length of a word
vector, also the frequency with which a word occurs plays an important
role.

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new tokens into the corpus with varying frequencies
and varying levels of noise in their co-occurrence distributions.  By
modelling the frequency and co-occurrence distributions of these tokens
on existing words in the corpus, we are able to study their effect on
word vectors independently of one another.  We can thus study a family
of tokens that all appear in the same context, but with different
frequencies, or study a family of tokens that all have the same
frequency, but appear in a different number of contexts.  Starting from
the limited number of contexts in which a word appears in the original
corpus, we can increase this number by inspersing the word in arbitrary
contexts at random.  The word thus looses its significance in a
controlled way.  Although we present our approach using the word2vec
CBOW model, these and related experiments could equally well be carried
out for other word embedding methods such as the word2vec skip-gram
model \cite{DistRepns,EfficientEstimation}, GloVe
\cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  We show furthermore that the
direction of the word vectors is independent of both word frequency and
co-occurrence noise.

This paper is structured as follows.  Section \ref{related-work} draws
connections to related work, while Section \ref{corpus-and-model} describes
the corpus and the CBOW model used in our experiments.
Section \ref{WFVE} describes a controlled experiment for varying word
frequency while holding the co-occurrence distribution fixed.
Section \ref{CNVE}, in a complementary fashion, describes a controlled
experiment for varying the level of noise in the co-occurrence
distribution of a word while holding the word frequency fixed.  The
final section, Section \ref{future-directions}, considers further questions
and possible future directions.
\end{section}

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}, who also based their work on word co-occurrences.
They define contextual distinctiveness of a target word $w$ as the
Kullback-Leibler divergence $\mathrm{D} \big(\p(w') ||\p(w'|w) \big)$
between the distribution of context words, $\p(w')$, and the
co-occurrence distributions of that target word $w$.  Specifically,
\begin{equation} 
\label{cd}
\mathrm{D} \big(\p(w') ||\p(w'|w) \big) = \sum_{w'} \p(w'|w) \log_2
\frac{\p(w'|w)}{\p(w')},
\end{equation} 
where the sum is over the set of \textcolor{blue}{content?} words $w'$.  They set the
window size to five words before and five words after the target word.
To reduce the computational burden, they recorded co-occurrences of 690
target words, randomly chosen from equally divided frequency bands, with
only the $n=500$ most frequent content words.  As pointed out by the
authors, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More specifically, if a
word $w$ tends to appear in a wide variety of contexts, the distribution
of the words it occurs with, i.e., the posterior distribution $\p(w'|w)$,
tends to diverge less from the prior distribution $\p(w')$, and its
contextual distinctiveness is accordingly small.  If, on the other hand,
$w$ typically appears in only a small number of different contexts, the
posterior distribution tends to diverge more from the prior, and a
larger value of contextual distinctiveness results.  As for
word significance, contextual distinctiveness summarizes a word's
contextual behavior in a single number. \textcolor{blue}{shorten?}

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  In that paper, which is also based
on word co-occurrence, the authors study adjective-noun composites.
They built a vocabulary from the 8k most frequent nouns and 4k most
frequent adjectives in a large general language corpus and added 22k
adjective-noun composites.  For each item in the vocabulary, they
recorded their co-occurrence within the same sentence with any of the
top 10k most frequent content words (nouns, adjectives or verbs) in the
corpus, and contructed word embeddings (of dimension 300) through
singular value decomposition of the co-occurrence matrix.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two
simplest being adding and component-wise multipying the adjective and
noun vectors.  They hypothesized that the length of the vectors thus
constructed can be used to distinguish acceptable and semantically
deviant adjective-noun composites.  Using a few hundred adjective-noun
composites selected by humans for evaluation, they found that deviant
composites have a shorter vector than acceptable ones, in accordance
with their expectation.

@@

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia datadump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word- (or ``term-'') frequency $tf$ is described in overview
in Table~\ref{word-occurrences-table}.  This base corpus is then
modified as described in Sections \ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

\begin{subsection}{Word2vec}\label{word2vec}
Word2vec, a feedforward neural network with a single hidden
layer, learns word vectors from word co-occurrences in an unsupervised
manner.  Word2vec comes in two versions.  In the continuous bag-of-words
(CBOW) model, the words appearing around a target word serve as input.
That input is projected linearly onto the hidden layer and the networks
then attempts to predict the target word on output.  Training is
achieved through backpropagation.  The word vectors are encoded in the
weights of the first synaptic layer, ``syn0''.  The weights of the
second synaptic layer (``syn1neg'', in the case of negative sampling) are typically discarded.  In the other model,
called skip-gram, target and context words swap places, so that the
target word now serves as input, while the network attempts to predict
the context words on output.

For simplicity only the word2vec CBOW word embedding with a single set
of hyperparameters is considered.  Specifically, a CBOW model with a
hidden layer of size 100 is trained using negative sampling with 5
negative samples, a window size of $10$, a minimum word occurrence of
$128$, and $10$ passes through the corpus.  Sub-sampling was not used so
that the influence of word frequency could be more clearly discerned.
Similar experimental results were obtained using hierarchical softmax,
but these are omitted for succinctness.  The relatively high
low-frequency cut-off is chosen to ensure that word vectors, in all but
degenerate cases, receive a sufficient number of gradient updates to be
meaningful.  This frequency cut-off results in a vocabulary of $81117$
words (only unigrams were considered).

The most recent revision of word2vec was used.\footnote{SVN revision 42,
  see \url{http://word2vec.googlecode.com/svn/trunk/}} The source code
for performing the experiments is made available on
GitHub.\footnote{\url{https://github.com/benjaminwilson/word2vec-norm-experiments}}

\begin{table}
	\input{words-occurrences.tex}
	\caption{ Number of words, by frequency band, as observed in the
          unmodified corpus.  }
	\label{word-occurrences-table}
\end{table}
\end{subsection}

\begin{subsection}{Replacement procedure}\label{replacement-procedure}
In the experiments detailed below, we modify the corpus in a controlled
manner by introducing tokens into the corpus via a replacement procedure
as follows.  Consider a word, say \word{cat}.  For each occurrence of
this word, a sample $i$, $1 \leqslant i \leqslant n$ is drawn from a
truncated geometric distribution, and that occurrence of the word
\word{cat} is replaced with the token \word{CAT\_i}.  In this way, the
word \word{cat} is replaced throughout the corpus by a family of tokens
with varying frequencies but approximately the same co-occurrence
distribution as \word{cat}.  That is, all these tokens are used in
roughly the same contexts as the original word.

The geometric distribution is truncated in order to limit the number of
tokens inserted into the corpus.  For any choice $0 < p < 1$ and
maximum value $n > 0$, the truncated geometric distribution is given by
the probability density function
\begin{equation} 
\label{distro}
P_{p, n} (i) = \frac{p^{i-1}  (1-p)}{1 - p^n}, \qquad 1
\leqslant i \leqslant n.
\end{equation} 
The factor in the denominator, which tends to unity in the limit $n \to
\infty$, assures proper normalisation.  We have chosen this distribution
because the probabilities decay exponentially base $p$ as a function of
$i$.  Of course, other distributions might equally well have been chosen
for the experiments.
\end{subsection}
\end{section}

\begin{section}{Varying word frequency}\label{WFVE}
In this first experiment, we investigate the effect of word frequency on
the word embedding.  Using the replacement procedure, we introduce a
small number of families of tokens into the corpus.  The tokens in each
family vary in frequency but, replacing a single word, all share a
common co-occurrence distribution.  This allows us to study the role of
word frequency in isolation, everything else being kept equal.  We
consider two types of tokens.

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
%
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
tokens do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table~\ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section ~\ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
number of occurrences sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum occurrence count of any token in
the corpus.  Figure~\ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Term frequency $tf$ of words chosen for the word
          frequency experiment. }
	\label{word-frequency-counts}
\end{table}

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section ~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with tokens \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
%
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we insperse the word uniformly at random throughout the corpus so
that its global frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the global word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher global frequency of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of iserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure ~\ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a global frequency of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

\subsection{Experimental results}
We next present the results of the word frequency experiment. We
consider the effect of word frequency on the direction and on the length
of word vectors separately.

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
%
Figure~\ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the tokens used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different tokens
associated with an experiment word is the number of times that its
frequency can be halved and remain above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and
so a limited number of samples is required to capture their
co-occurrence distributions.  We therefore expect that the vectors
representing the tokens derived from any of these words can be properly
learned from the modified corpus, even though their combined number of
occurences equals the term frequency of that word in the original
corpus.  The red blocks in Figure ~\ref{word-frequency-experiment-heatmap}
clearly demonstrate that the direction of these vectors is constant for
each family of tokens.  In other words, the direction of these vectors
is only determined by the original word and is independent of how often
that words appear, provided sufficient samples have been processed for
the word vector to settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the tokens occur increasingly less often.  This is consistent with
Figure ~\ref{word-frequency-experiment-heatmap}, where it is apparent that
the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the tokens
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these tokens.
Table~\ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each token \word{THE\_i}, supports this further.  In the
discussions below, the tokens \word{VOID\_i} and
\word{THE\_i} whose vectors have a cosine similarity less than $0.8$
with \word{VOID\_1} and \word{THE\_1}, respectively are therefore
discarded.

From these results, we conclude that word frequency has no effect on the
direction of a word vector, given a sufficient number of samples to
learn the co-occurrence distribution of the word.
%For well-learned representations, word frequency has no effect on word
%vector direction.

\subsubsection{Word frequency and vector length}
%
We next consider the effect of frequency on word vector length.
Figure~\ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the number of occurences of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure ~\ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e., the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure ~\ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $tf=10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless tokens
\word{VOID\_i} are approximately constant.  Since we already found the
direction to be also constant, it is sensible to speak of the word
vector of \word{VOID} irrespective of its frequency.  In particular, the
token \word{VOID\_1} may be taken as an approximation.

\begin{figure}
	\includegraphics[scale=0.5]{word-frequency-experiment-heatmap}
	\caption{ Heatmap of the cosine similarity of the vectors
          representing some of the tokens used in the word frequency
          experiment.  The words other than \word{the} and \word{VOID}
          were chosen randomly.  }
	\label{word-frequency-experiment-heatmap}
\end{figure}

\begin{table}
	\input{cosine-similarity.tex}
	\caption{ Words in the original vocabulary most similar to the
          tokens \word{THE\_i}, and their cosine similarity with the
          most frequent such token, \word{THE\_1}.  It is apparent from
          the nearest neigbour list that the vectors of the
          low-frequency tokens have not been adequately trained.  }
	\label{word-frequency-experiment-cosine-similarity}
\end{table}
